---
title: "Fisher and t-tests"
author: "Angelina Legostaeva"
date: "2025-02-28"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


1. hist
2.qqnorm
3. Fisher
4. computational method to estimate the variance of the MLE for λ
5. 95% Wald confidence interval for λ

6. wald
7. p-value

8. the computational approach and optim()
9. the estimated variance matrix for your MLEs

10. t-stat
11 t.test




1. 
Data:
```{r}
expo <- scan("expo.txt")
hist(expo, breaks = seq(0, 20, 1), main = "Expo Histogram")
qqnorm(expo)
summary(expo)
```
###################################################################################

(a) Use observed Fisher information to estimate the variance of 
the MLE for the rate parameter λ in the Exponential distribution, ˆV (ˆλ).

```{r}
n <- length(expo)

lambda_exp <- 1 / mean(expo)
lambda_exp

var_exp <- lambda_exp^2 / n
var_exp

fisher <- n / lambda_exp^2
fisher
```

(b)  No, the formulas are different. 
```{r}
n <- length(expo)

lambda_poi <- mean(expo)
lambda_poi

var_poi <- lambda_poi / n
var_poi
```
###################################################################################

(c) Use computational method to estimate the variance of the MLE for λ, ˆV (ˆλ).
```{r}
set.seed(352) 
x <- rexp(n, rate = 0.5) 

nll_exp <- function(lambda, data) {
  n <- length(data)
  return(- (n * log(lambda) - lambda * sum(data))) 
}

expo.mle <- optim(par = 1, fn = nll_exp, data = expo, method = "Brent", lower = 0.001, upper = 10, hessian = TRUE)
expo.mle$par
expo.mle$hessian
```
###################################################################################

(d) Construct a 95% Wald confidence interval for λ. Interpret your confidence interval and
make a conclusion about the true value of λ being 0.5.
```{r}
z <- 1.96

expo.var <- solve(expo.mle$hessian)
expo.var

se <- sqrt(diag(expo.var))

lower <- expo.mle$par - z * se
upper <- expo.mle$par + z * se

c(lower, upper)
```
###################################################################################

(e)
P-value is 0.00117656 < 0.5, so our p-value is small, while the sample size is big. 
So, it would be unusual if the alternative hypothesis would be true. 
There is evidence for the alternative hypothesis. Reject H0 : λ = 0.5.
```{r}
W <- (expo.mle$par - 0.5)^2 / (1 / expo.mle$hessian)
W

# P-value
1 - pchisq(W, df = 1)
```
###################################################################################

2. file womensheights.txt and Normal distribution

```{r}
height <- scan("womensheights.txt")
summary(height)
```

(a) Using data and the computational approach and optim(),
```{r}
mu_ <- mean(height)
sigma2_ <- var(height)

h.nll <- function(params) {
  mu <- params[1]
  sigma2 <- params[2]
  n <- length(height)
  nll <- (n/2) * log(2 * pi * sigma2) + sum((height - mu)^2) / (2 * sigma2)
  return(nll)
}

optim_mle <- optim(par = c(mu_, sigma2_), fn = h.nll, hessian = TRUE)

mle_mu <- optim_mle$par[1] 
mle_mu

mle_sigma2 <- optim_mle$par[2]
mle_sigma2
```
###################################################################################

i. find the MLEs for μ and σ2;
```{r}
mu <- 162.05621  
sigma2 <- 52.9174
```
ii. find and state the estimated variance matrix for your MLEs;
```{r}
matrix.var <- solve(optim_mle$hessian)
matrix.var
```

iii. find a 99% Wald confidence interval for μ, the average height of American adult women;
```{r}
se_mu <- sqrt(matrix.var[1, 1])

# 1 - ((1-0.99)/2) = 0.995
mu_ci <- optim_mle$par[1] + c(-1, 1) * qnorm(0.995) * se_mu
mu_ci
```
###################################################################################

iv. find a 99% Wald confidence interval for σ2, the variance of American adult women.
```{r}
se_sigma2 <- sqrt(matrix.var[2, 2])

sigma2_ci <- optim_mle$par[2] + c(-1, 1) * qnorm(0.995) * se_sigma2
sigma2_ci
```


(b) How would you find a 99% confidence interval for σ, the standard deviation of heights
of American adult women?
```{r}
#sigma = sqrt(sigma^2)
sigma_ci <- sqrt(sigma2_ci)
sigma_ci
```
###################################################################################

(c) mu = 161.3 
Wald test:
P -value is 1.776946e-10 which is much smaller than 0.5, 
so I reject the null hypothesis H0 that 
the average height of American adult women is 161.3 cm. 
```{r}
w <- (optim_mle$par[1] - 161.3)^2 / matrix.var[1, 1]
w

1 - pchisq(w, df = 1)
```

###################################################################################

```{r}
t_stat <- (mle_mu - 161.3) / se_mu
t_stat
2 * (1 - pnorm(abs(t_stat)))
```
###################################################################################

(d) Use t.test to perform a t test for the above hypothesis as well as produce a 99%
confidence interval. Yes, the same conclusion, because the p-values is still really small.
```{r}
t <- t.test(height, mu = 161.3, conf.level = 0.99)
t
```


(e) Using data and the theory approach,

i. find the MLEs for μ and σ2;
```{r}
n <- length(height)
      
theory.mu <- mean(height)
theory.sigma2 <- var(height) * (n - 1) / length(height)

theory.mu
theory.sigma2
```

ii. find a 99% confidence interval for μ, the average height of American adult women;
```{r}
n <- length(height)
se.theory <- sqrt(theory.sigma2 / n)

lower_mu <- theory.mu - qnorm(0.995) * se.theory
upper_mu <- theory.mu + qnorm(0.995) * se.theory
c(lower_mu, upper_mu)
```

iii. find a 99% confidence interval for σ2, the variance of heights of American adult
women.

```{r}
n <- length(height)
se.sd <-  sqrt(2 * theory.sigma2) / sqrt(n)

lower_mu <- theory.sigma2 - qnorm(0.995) * se.sd
upper_mu <- theory.sigma2 + qnorm(0.995) * se.sd
c(lower_mu, upper_mu)
```

iv. mu = 5.3
P-value is much smaller than 0.05 (but big data set); the observed value is different from 5.3, 
So, we would reject the null hypothesis H0: mu = 5.3. Conclusion, I used the Wald test for 
the mle estimate for the sigma or the standard deviation to approve (or disprove) 
the hypothesis about the standard deviation of women’s heights is, but in Belgium  if it’s 5.3 cm.
```{r}
v <- 2 * 5.3^4 / 3766
w <- (theory.sigma2 - 5.3^2)^2 / v
1 - pchisq(w, df = 1)
```
