---
title: "Functions"
author: "Angelina Legostaeva"
date: "2025-04-09"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
roots <- scan("roots.txt")
```

Negative binomial:
```{r}
nbinom.nll <- function(params, data){
  size <- params[1]
  prob <- params[2]
  lik <- dnbinom(data, size, prob)
  loglik <- log(lik)
  nll <- -sum(loglik)
  return(nll)
}
```
```{r}
nbinom.nll(c(1, 0.5), data = roots)
```

Now use optim():
```{r}
roots.nb.mle <- optim(par = c(r.mom, p.mom), fn = nbinom.nll, data = roots)
roots.nb.mle
```

AIC NB:
```{r}
nb.aic <- 2 * roots.nb.mle$value + 2 * 2
nb.aic
```

##############################################################################
Poisson:

98% profile likelihood confidence interval for λ: 2.318556 to 2.696201
```{r}
poi <- function(lambda){
  loglik <- sum(dpois(laliga, lambda, log = TRUE))
  return(-loglik)
}

control.list = list(label = "lambda", est = lambda, low = 0, upp = 10)
library(Bhat)
plkhci(control.list, poi, "lambda", prob = 0.98)
```

```{r}
lambda.hat <- mean(roots)
lambda.hat
var(roots)
p <- dpois(0:9, lambda.hat)
round(40 * p, 2)
```

Find the AIC:
```{r}
pois.ll <- sum(log(dpois(roots, lambda.hat)))
-2 * pois.ll + 2 * 1
```


Wald Ci;
```{r}
lambda.hat <- mean(counts)
n <- length(counts)

# Standard error
se.lambda <- sqrt(lambda.hat / n)

# Wald 95% CI
ci.lambda <- lambda.hat + qnorm(0.975) * se.lambda

ci.lambda
```

#############################################################################

Zero-inflated Poisson NLL function:
```{r}
pois0.nll <- function(pars, data){
  p <- pars[1]
  lambda <- pars[2]
  pois.lik <- dpois(data, lambda)
  real.lik <- rep(NA, length(data))
  which0 <- (data == 0)
  real.lik[which0 == FALSE] <- (1 - p) * pois.lik[which0 == FALSE]
  real.lik[which0 == TRUE] <- p + (1 - p) * pois.lik[which0 == TRUE]
  loglik <- log(real.lik)
  return(-sum(loglik))
}
```

```{r}
roots.zip.mle <- optim(par = c(19/40, 2.45), fn = pois0.nll, data = roots)
roots.zip.mle
```


Get MOMs:
```{r}
x.bar <- mean(roots)
sigma2.hat <- mean(roots^2) - x.bar^2
p.mom <- x.bar / sigma2.hat
r.mom <- p.mom / (1 - p.mom) * x.bar
c(r.mom, p.mom)
```

Now use optim():
```{r}
roots.nb.mle <- optim(par = c(r.mom, p.mom), fn = nbinom.nll, data = roots)
roots.nb.mle
```

```{r}
roots.zip.mle <- optim(par = c(19/40, 2.45), fn = pois0.nll, data = roots)
roots.zip.mle
```

or
```{r}
art.zip <- optim(par = c(mean(biochemists$art == 0), mean(biochemists$art)), fn = pois0.nll, data = biochemists$art)
art.zip$par
```



AIC Zero-inflated:
```{r}
2 * roots.zip.mle$value + 2 * 2
```

#############################################################################


Gamma:
```{r}
mean.wwe <- mean(wwe)
lambda.init <- 1 / mean.wwe
k.init <- 1
```

```{r}
x.bar <- mean(WWE)
sigma2.hat <- mean(WWE^2) - x.bar^2
# Fit a gamma
gamma.nll <- function(pars, data){
  k <- pars[1]
  lambda <- pars[2]
  lik <- dgamma(data, k, lambda)
  return(-sum(log(lik)))
}

k.mom <- x.bar^2 / sigma2.hat
lambda.mom <- x.bar / sigma2.hat
gamma.mle <- optim(par = c(k.mom, lambda.mom), fn = gamma.nll, data = WWE, hessian = TRUE)
```

Wald test for H0 : k = 1
```{r}
nerve.var <- solve(nerve.mle$hessian)
w <- (nerve.mle$par[1] - 1)^2 / nerve.var[1, 1]
w

#p-value
1 - pchisq(w, df = 1)
```

or

```{r}
mle <- nerve.mle$par
vcov.mat <- solve(nerve.mle$hessian)
se <- sqrt(diag(vcov.mat))


ci.k <- mle[1] + qnorm(0.975) * se[1]
ci.lambda <- mle[2] + qnorm(0.975) * se[2]
```


Gamma AIC:
```{r}
2 * gamma.mle$value + 2 * 2
```

##################################################################################

```{r}
binom.nloglik <- function(p){
  lik <- dbinom(10, 100, p)
  return(-log(lik))
}

library(Bhat)
control.list <- list(label = "p", est = 0.1, low = 0, upp = 1)
plkhci(control.list, binom.nloglik, "p")
```


AIC Binom or Bern.:
```{r}
x <- 80  # number of successes
n <- 100  # number of trials
p.hat <- x / n  # Estimated probability p

loglik_binom <- dbinom(x, n, p.hat, log = TRUE)

k <- 1  # Binomial has one parameter (p)
AIC_binom <- 2 * k - 2 * loglik_binom
AIC_binom
```


#########################################################################################################################################

Geometric data/distribution, 95% CI for p 

use plkhci():
```{r}
geom.nloglik2 <- function(p){
  nloglik1 <- -3 * log(1 - p) - log(p)
  nloglik2 <- -2 * log(1 - p) - log(p)
  nloglik <- nloglik1 + nloglik2
  return(nloglik)
}

library(Bhat)
control.list <- list(label = "p", est = 2/7, low = 0, upp = 1)
plkhci(control.list, geom.nloglik2, "p")
```

#########################################################################################################################################

Binomial:
```{r}
binom.nll <- function(p){
  lik <- dbinom(binomdata, size = 100, prob = p) # a vector
  loglik <- log(lik) # a vector
  nloglik <- -sum(loglik) # a number
  return(nloglik)
}

# Check the function works:
binom.nll(0.7)
```

```{r}
p.hat <- 12614 / 38663  # example for digit 1
n <- 38663
se <- sqrt(p.hat * (1 - p.hat) / n)

ci.p <- p.hat + qnorm(0.975)* se
```

############################################################################

Observe x = 5. What's the likelihood?

```{r}
mu <- seq(0, 10, 0.001)
lik <- dnorm(5, mean = mu, sd = 1)
plot(mu, lik, type = "l")
```

Write an NLL function:

```{r}
normal.nll <- function(mu){
  lik <- dnorm(5, mu, 1)
  nll <- -log(lik)
  return(nll)
}
```
```{r}
optimize(normal.nll, c(-10, 10))
```

##############################################################################

Exponential distribution CDF and PDF:
```{r}
curve(pexp(x, rate = 1), from = 0, to = 10)
curve(dexp(x, rate = 0.4), from = 0, to = 10)
```


MLE for the rate parameter is 1/sample mean.
```{r}
mean(snoq)
lambda.mle <- 1 / mean(snoq)
lambda.mle
```


Exponential AIC:
```{r}
-2*sum(log(dexp(WWE, 1/mean(WWE)))) + 2*1
```


Wald CI:
```{r}
lambda.mle <- 1 / mean(wwe)

n <- length(wwe)
se.lambda <- lambda.mle / sqrt(n)

ci.lambda <- lambda.mle + qnorm(0.975) * se.lambda
```

##############################################################################

Uniform Model


Write a function to find the MLE of the maximum of a Uniform distribution from simulated data:

```{r}
uniMLE <- function(n, b){
  data <- runif(n, min = 0, max = b)
  b.mle <- max(data)
  return(b.mle)
}
```
```{r}
uniMLE(n = 5, b = 100)
uniMLE(n = 100, b = 100)
```


##############################################################################

Normal distribution: estimators for the variance 

Do a simulation to compare the biases of the MLE for variance and the sample variance:
```{r}
norm.sim <- function(n){
  data <- rnorm(n) # True parameter value: variance = 1
  sigma2 <- mean((data - mean(data))^2) # MLE
  s2 <- var(data) # Sample variance
  return(c(sigma2, s2))
}
```


##############################################################################

Uniform distribution: MLE for the upper bound b

Write a function to find the MLE of the upper bound of a Uniform distribution from simulated data:

```{r}
uniMLE <- function(n, b){
  data <- runif(n, min = 0, max = b) # simulate data from U(0,b)
  b.mle <- max(data) # compute MLE of b from simulated data
  return(b.mle)
}

uniMLE(n = 20, b = 10)
uniMLE(n = 50, b = 10)
uniMLE(n = 50, b = 80)
```

##############################################################################
G-statistic:
```{r}
lambda.hat <- mean(TotalGoals)
lambda.hat

p <- dpois(0:6, lambda.hat)

exp <- n * p
expected <- c(exp, n - sum(exp))

observed <- c(22, 55, 99, 84, 65, 32, 16, 7)
  
G <- 2 * sum(observed * log(observed / expected))
G

p_value <- 1 - pchisq(G, 6)
p_value
```

```{r}
obs <- c(12614, 6443, 4563, 3581, 2951, 2533, 2227, 1988, 1763) * 1000
total <- sum(obs)
f <- c(0.3010, 0.1761, 0.1249, 0.0969, 0.0792, 0.0669, 0.0580, 0.0512, 0.0458)
expected <- total * f

loglik_multinomial <- sum(obs * log(obs / expected))

k <- length(f)  # Number of categories (9 for Benford)
AIC_multinomial <- 2 * k - 2 * loglik_multinomial
AIC_multinomial
```


##############################################################################

BICs for our probability models:

```{r}
# Poisson
-2 * sum(log(dpois(roots, mean(roots)))) + 1 * log(40)
# Negative binomial
2 * 81 + 2 * log(40)
# Zero inflated Poisson
2 * roots.zip.mle$value + 2 * log(40)
```


##############################################################################