---
title: "G-test, LL ratio, plkchi()"
author: "Angelina Legostaeva"
date: "2025-03-25"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


####################################################################################

1. G-test
2. Likelihood ratio test

3. profile likelihood confidence interval for λ
poi <- function(lambda){}

control.list 
plkhci(control.list, poi, "lambda", prob = 0.98)

4. wlad ci plkhci

5. AIC not the easy way

####################################################################################
1.


```{r}
observed <- c(29, 19, 18, 25, 17, 10, 15, 11)
n <- sum(observed)
expected <- rep(n/8, 8)

G <- 2 * sum(observed * log(observed / expected))
1 - pchisq(G, df = 8 - 1)
```
Conclusion: P-value is 0.02, which is smaller than 0.05.
I am rejecting the null hypothesis that a horse’s starting position does not affect its chance of winning.
So, for the alternative hypothesis: a horse’s starting position affect its chance of winning.

2. 
Observed:
• Tall cut-leaf: 926
• Tall potato-leaf: 288
• Dwarf cut-leaf: 293
• Dwarf potato-leaf: 104


```{r}
observed <- c(926, 288, 293, 104)
n <- sum(observed)
expected <- n * c(9/16, 3/16, 3/16, 1/16)

G <- 2 * sum(observed * log(observed / expected))
1 - pchisq(G, df = 4 - 1)
```
Conclusion: p-value is 0.7, which larger than 0.05.
So, I am failing to reject the null hypothesis that the observations are consistent with the expected proportions.

####################################################################################

3. The file laliga.txt contains the number of goals scored in each of the 380 games of the
2021–22 Spanish La Liga soccer season (between 0 and 8.) We wish to model this data using
a Poisson distribution with parameter λ.

Data: 
```{r}
laliga <- scan("laliga.txt")
summary(laliga)
```


(a) The numerical value of the MLE for λ is 2.502632.
```{r}
lambda <- mean(laliga)
lambda
```

(b) The variance of ˆλ = λ / n
The variance of ˆλ is 0.006585873
```{r}
n <- length(laliga)

var <- lambda / n
var
```

(c) H0:lambda = 2.5
Perform a likelihood ratio test of this hypothesis, giving a test statistic, P-value, and conclusion.
```{r}
W <- (lambda - 2.5)^2 / var
W
p_value3 <- 1 - pchisq(W, df = 1)
p_value3
```
####################################################################################

Option 1: Likelihood ratio test.
```{r}
# Unrestricted loglik
loglik.max <- sum(log(dpois(laliga, mean(laliga))))

# Restricted loglik
loglik.null <- sum(log(dpois(laliga, 2.5)))

# LR stat
G <- 2 * (loglik.max - loglik.null)

# P-value
1 - pchisq(G, df = 1)
```

Conclusion: The P-value is about 0.97, which is greater than 0.05. 
This may be an underestimate, so there's no current evidence of a difference.
I an failing to reject the null hypothesis that lambda = 2.5.
####################################################################################


(d) 
98% profile likelihood confidence interval for λ: 2.318556 to 2.696201
```{r}
poi <- function(lambda){
  loglik <- sum(dpois(laliga, lambda, log = TRUE))
  return(-loglik)
}

control.list = list(label = "lambda", est = lambda, low = 0, upp = 10)
library(Bhat)
plkhci(control.list, poi, "lambda", prob = 0.98)
```

98% t-confidence interval for λ: 2.296151 to 2.709112
```{r}
t.test(laliga,mu=2.5,conf.level=0.98)
```

Conclusion:yes, the results are really similar. 
For 98% profile likelihood confidence interval, λ is from 2.3 to 2.7.
For 98% t-confidence interval, λ is from 2.31 to 2.7. 

####################################################################################

```{r}
# Wald
x.bar <- mean(laliga)
se <- sqrt(x.bar / 380)
wald.ci <- x.bar + qnorm(c(.01, .99)) * se
# (2.313841, 2.691423)
# Profile
laliga.nll <- function(lambda){
lik <- dpois(laliga, lambda)
return(-sum(log(lik)))
}
library(Bhat)
control.list <- list(label = "lambda", est = x.bar, low = 0, upp = 99)
profile.ci <- plkhci(control.list, laliga.nll, "lambda", prob = 0.98)
# (2.318555, 2.696199
```
####################################################################################


4. 
Data:
```{r}
hurricanes <- read.csv("hurricanes(2).csv")
summary(hurricanes$alldeaths)
alld <- hurricanes$alldeaths
```
Poisson: 4358.845
```{r}
lambda.hat <- mean(alld)

pois.ll <- sum(log(dpois(alld, lambda.hat)))
poi_aic <- -2 * pois.ll + 2 * 1
poi_aic
```

Negative Binomial: 708.6559
```{r}
nbinom.nll <- function(params, data){
  size <- params[1]
  prob <- params[2]
  lik <- dnbinom(data, size, prob)
  loglik <- log(lik)
  nll <- -sum(log(lik))
  return(nll)
}

x.bar <- mean(alld)
sigma2.hat <- mean(alld^2) - x.bar^2

p.mom <- x.bar / sigma2.hat
r.mom <- p.mom / (1 - p.mom) * x.bar

nb.mle <- optim(par = c(r.mom, p.mom), fn = nbinom.nll, data = alld, hessian = TRUE)
nb.mle

nb_aic <- 2 * nb.mle$value + 2 * 2
nb_aic
```

Zero-Inflated Poisson:3986.837

```{r}
pois0.nll <- function(pars, data){
  p <- pars[1]
  lambda <- pars[2]
  pois.lik <- dpois(data, lambda)
  real.lik <- rep(NA, length(data))
  which0 <- (data == 0)
  real.lik[which0 == FALSE] <- (1 - p) * pois.lik[which0 == FALSE]
  real.lik[which0 == TRUE] <- p + (1 - p) * pois.lik[which0 == TRUE]
  loglik <- log(real.lik)
  return(-sum(loglik))
}

p.init <- mean(alld == 0)
lambda.init <- mean(alld)
c(p.init, lambda.init)

poi0.mle <- optim(par = c(0.109, 20.652), fn = pois0.nll, data = alld, hessian = TRUE)
poi0.mle

pois0.aic <- 2 * poi0.mle$value + 2 * 2
pois0.aic
```


Zero-Inflated Negative Binomial: 697.3579

```{r}
nll.b <- function(pars){
  p <- pars[1]
  lambda <- pars[2]
  prob <- pars[3]
  pois.lik <- dnbinom(alld, lambda, prob) 
  real.lik <- rep(NA, length(alld))
  which0 <- (alld == 0) 
  real.lik[which0 == FALSE] <- (1 - p) * pois.lik[which0 == FALSE] 
  real.lik[which0 == TRUE] <- p + (1 - p) * pois.lik[which0 == TRUE]
  loglik <- log(real.lik)
  return(-sum(loglik))
}

nll.b.mle <- optim(par = c(0.5, 1, 0.5), fn = nll.b, hessian = TRUE)
nll.b.mle

nb0.aic <- 2 * nll.b.mle$value + 2 * 3
nb0.aic
```

I will choose Negative Binomial (708.6559), because it has the smallest AIC.
Not the Zero-Inflated Negative Binomial (697.3579), because it is an impossible model due to negative $par value (-1.39666082).
