---
title: "Problem Set 11"
author: "Angelina Legostaeva"
date: "2025-04-29"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



#######################################################################################

1. Logistic Regression
2. glm, family = binomial
3. Divide-by-4 rule
4. logit(p) and prob. scale
5. model’s probabilities for each of the four largest - type = "response"

6. data.frame
7. plot

8. Poisson Regression
9. setting terms
10. 1 - sum(dpois(0:9, 10.15112)) #probability

#######################################################################################



1 Logistic Regression

1. Data:
```{r}
elec <- read.table("elec16.txt", header = T)
elec$vote <- (elec$vote == "Clinton")
#Clinton
```


2. (a) regression equation = 0.04410 + 0.03042 * education
```{r}
vote_lr <- glm(vote ~ education, family = binomial, data = elec)
summary(vote_lr)
```
#######################################################################################

(b) According to the divide-by-4 rule, the model probability of voting for Clinton
increases by 0.76%
```{r}
db4 <- 0.03042/4
db4
```

#######################################################################################

(c) Find the model’s probabilities for voting for Clinton for each of the six levels of education
the long way (i.e., first get the value of logit(p) and then use the inverse of logit to get
the probability).

```{r}
logit.p <- 0.04410 + 0.03042 * (1:6)
logit.p

# Probability scale
exp(logit.p) / (1 + exp(logit.p))
```


#######################################################################################


3. Now, we want to fit a logistic regression to predict the probability of vote using race as a
categorical x-variable.

(a) 
Largest group is White (7704), then Black (962), then Hispanic (635), and Asian(272).
Smallest is Middle Eastern which is only 16, so not really reliable. 
However, other samples have good sizes.
```{r}
table(elec$race)
```


(b) Fit the logistic regression and show the coefficients.
```{r}
race_lr <- glm(vote ~ race, family = binomial, data = elec)
summary(race_lr)
```

(c) The baseline group is Asian. 
The highest probability of voting for Clinton is Black.
The lowest probability is Other.

#######################################################################################


(d) Find the model’s probabilities for voting for Clinton for each of the four largest groups
(in terms of the sample size), using the predict() function.
```{r}
predict(race_lr, 
        newdata = data.frame(race = c("White", "Black", "Hispanic", "Asian")), 
        type = "response")
```


4. Fit a logistic regression to predict the probability of vote using race and education (separately or with no interaction) as x-variables.
```{r}
vote_race <- glm(vote ~ race + education, family = binomial, data = elec)
summary(vote_race)
```

#######################################################################################################################

(a) 
White:
```{r}
white <- predict(vote_race, newdata = data.frame(race = "White", education = 1:6), type = "response")
white
```
Black:
```{r}
black <- predict(vote_race, newdata = data.frame(race = "Black", education = 1:6), type = "response")
black
```
Hispanic:
```{r}
hisp <- predict(vote_race, newdata = data.frame(race = "Hispanic", education = 1:6), type = "response")
hisp
```

(b) Draw a graph of the results (my recommendation would be to do a line for White, a line
for Black, and a line for Hispanic on the same axes, with education on the x-axis.)

```{r}
plot(1:6, white, type = "l",
     xlab = "Education", ylab = "Probability",
     main = "White = blue; Black = red; Hispanic = green",
     ylim = c(0, 1), col = "blue")

lines(1:6, black, col = "red")
lines(1:6, hisp, col = "green")
```
#######################################################################################################################

5. Fit a logistic regression to predict the probability of vote using race, education, and their
interaction as x-variables.
```{r}
race_ed <- glm(vote ~ race * education, family = binomial, data = elec)
#summary(race_ed)
```

(a) Find the model probability of voting for Clinton for each combination of race being
White, Black, or Hispanic with each education level from 1 to 6 (for a total of 18
predictions.) Draw a graph of the results.
```{r}
white_2 <- predict(race_ed, newdata = data.frame(race = "White", education = 1:6), type = "response")
black_2 <- predict(race_ed, newdata = data.frame(race = "Black", education = 1:6), type = "response")
hisp_2 <- predict(race_ed, newdata = data.frame(race = "Hispanic", education = 1:6), type = "response")
white_2
black_2
hisp_2
```
```{r}
plot(1:6, white_2, type = "l",
     xlab = "Education", ylab = "Probability",
     main = "White = blue; Black = red; Hispanic = green",
     ylim = c(0, 1), col = "blue")

lines(1:6, black_2, col = "red")
lines(1:6, hisp_2, col = "green")
```


(b) This graph's green line(Hispanic) is pointed downwards.



6. vote_race model has the lower BIC (12813.53)

```{r}
BIC(vote_race, race_ed)
```



7. Add religious.importance to the model with lower BIC from question 6 and fit one more
logistic regression.

(a) Yes, this logistic model have an even lower BIC?
```{r}
relig <- glm(vote ~ race + education + religious.importance, family = binomial, data = elec)
BIC(relig)
```


(b) Show the regression coefficients. Give a clear interpretation of what the coefficient for
religious.importance means.

For each one unit increase in religious.importance, voting decrease by 48%
```{r}
summary(relig)
```
```{r}
exp(-0.6543) - 1
```

#######################################################################################################################
#######################################################################################################################

2 Poisson Regression:

1. Fit a Poisson regression that models the log of expected absences as an additive function
of the race, gender, school, and learner variables (with no interactions), and give the
coefficients of the regression model.

```{r}
school <- read.table("schooldays.txt", header = T)

poisson <- glm(absent ~ race + sex + school + learner, family = poisson, data = school)
summary(poisson)
```
Poisson regression = 2.80601 -0.48843 * racenon-aboriginal + 0.15616 * sexmale -0.14596 * schoolF1 + 
0.30205 * schoolF2 + 0.20088 * schoolF3 + 0.04870 * learnerslow

2. Find the expected days absent (according to the model) for a non-aboriginal female child who
is an average learner at an F0 school.
```{r}
predict(poisson, newdata = data.frame(race = "non-aboriginal", sex = "female", school = "F0", learner = "average"), 
        type = "response")
```
3. Assuming the Poisson model is correct, what is the probability a non-aboriginal female child
who is an average learner at an F0 school is absent for at least ten days?

```{r}
1 - sum(dpois(0:9, 10.15112))
```

