---
title: "Pair Plots and Interactions"
author: "Angelina Legostaeva"
date: "2025-04-11"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Data:
```{r}
rust <- read.table("rust.txt", header = T)
```

Create a data frame 
Draw a pairs plot of the numeric variables

```{r}
DeathRate <- rust$Deaths / rust$Population * 100000
Density <- rust$Population / rust$Area
PctWhite <- rust$PctWhite
logIncome <- log(rust$Income)
Swing <- rust$Trump - (1 - rust$Obama)
County <- rust$County

rust2 <- data.frame(DeathRate, Density, PctWhite, logIncome, Swing, County)
rust2
```


```{r}
pairs(DeathRate ~ Density + PctWhite + logIncome + Swing, data = rust2, pch = 16)
```

no interaction = 22.72 + 150.9 * Swing + 0.005408 * Density + error
```{r}
no_int <- lm(DeathRate ~ Swing + Density, data = rust2)
no_int
```

DeathRate = (22.57 + 0.007068) * Density + (136.1 + 0.0534) * Density + error

DeathRate = 22.577 * Density + 136.1534 * Density + error
```{r}
q1model <- lm(DeathRate ~ Swing + Density + Swing:Density, data = rust2)
q1model
```
AIC:
```{r}
AIC(no_int)
AIC(q1model)
```
Model with interaction has lower AIC, so it is better.


(a) Density = 55 (similar to Randolph County, the lowest density county in Indiana)

DeathRate = 22.577 * 55 + 136.1534 * 55 + error


(b) Density = 369 (similar to Monroe County, IN)

DeathRate = 22.577 * 369 + 136.1534 * 369 + error


(c) Density = 2375 (similar to Marion County, IN, i.e. Indianapolis)

DeathRate = 22.577 * 2375 + 136.1534 * 2375 + error


For all ages:
```{r}
# Work out equations for Randolph, Monroe, and Marion Counties:

intercepts <- coef(q1model)[1] + coef(q1model)[3] * c(55, 369, 2375)

slopes <- coef(q1model)[2] + coef(q1model)[4] * c(55, 369, 2375)

intercepts


slopes
```


Plot:
```{r}
plot(rust2$Swing, rust2$DeathRate, main = "Blue: age 55; orange: age 369; green: age 2375" )

abline(22.96331, 139.0838, col = "blue")
abline(25.18274, 155.8740, col = "orange")
abline(39.36167, 263.1392, col = "green")
```


####################################################################################

The distribution does not look symmetric (not normally distributed), 
because it is curved and has some skewdness to the right.

This suggest about possible improvements to the model. For example, 
we can change some of the predictors and(or) delete some outlier points.
Another solution for right-skewdness is to make a transformation
to the data. For example, logarithms or sqrt for right skewdness.

```{r}
residuals <- residuals(q1model)

qqnorm(residuals, main = "Q-Q Plot for Residuals for q1model")
qqline(residuals, col = "blue")

hist(residuals, main = "Histogram for Residuals of q1model", 
     xlim = c(-40, 100), ylim = c(0, 80))
```

####################################################################################


(a) Forward selection starting from an empty model:
AIC: 1373.103

```{r}
full <- lm(DeathRate ~ Density + PctWhite + logIncome + Swing, data = rust)

empty <- lm(DeathRate ~ 1, data = rust)
empty

forw.empty <- step(empty, scope = formula(full), direction = "forward")
summary(forw.empty)

AIC(forw.empty)
```
(b) Backward selection starting from a full model with all predictors:


AIC: 1373.103

```{r}
full <- lm(DeathRate ~ Density + PctWhite + logIncome + Swing, data = rust)
full

back.full <- step(full, direction = "backward")
back.full
AIC(back.full)
```


(c) Both-ways stepwise selection starting from an empty model:

AIC: 1373.103

```{r}
both.empty <- step(empty, scope = formula(full), direction = "both")
both.empty
AIC(both.empty)
```



(d) Both-ways stepwise selection starting from a full model:

AIC: 1373.103

```{r}
both.full <- step(full, direction = "both")
both.full
AIC(both.full)
```
All these methods result in the same model: 22.72 + 0.005408 * Density + 150.9 * Swing + error

All these models share the same AIC as well: 1373.103
 